Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-06-25 analysis/paper/paper.qmd
updates of intro

At present, the technological category and function of spheroids remain central topics of debate. Broadly, spheroids have been interpreted as hammerstones ([@mussi2025; @schick1994], cores ([@sahnouni1997]), or projectiles ([@leakey1979]). Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]). Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; [\@mora2005]; [@clark1955]). In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]).

@@

We conducted quantitative morphological and technological analyses of the spheroids recovered from the QSY site. Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes. These analyses aimed to clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.

# Methods

>>>>

At present, the technological category and function of spheroids remain central topics of debate. Broadly, spheroids have been interpreted as hammerstones ([@mussi2025; @schick1994], cores ([@sahnouni1997]), or projectiles ([@leakey1979]). Closely tied to this debate is the question of their production strategy, which remains unclear. Experimental replication and site-specific diacritical analyses have produced divergent interpretations. One view suggests that spheroids emerged unintentionally as the by-products of core exhaustion, which represents a low-effort strategy aimed at flake production ([@sahnouni1997]). Another perspective holds that spheroids were naturally produced through high-intensity battering ([@schick1994]; \[\@mora2005\]; [@clark1955]). In contrast, a third view proposes that spheroids reflect a conceptual template and were produced through a deliberate shaping strategy ([@tittonSubspheroidsLithicAssemblage2020; @Muller2023; @texier2014]).

@@

We conducted quantitative morphological and technological analyses, including the novel spherical harmonic analysis, on the spheroids recovered from the QSY site. Our goal was to evaluate whether these spheroids exhibited standardized morphology and production patterns, and whether their tendency toward spherical form resulted from deliberate and systematic reduction processes. These analyses aimed to clarify the technological strategy behind spheroid production at QSY, within the broader context of Middle Pleistocene lithic practices in northern China.

# Context of QSY and its spheroids

The QSY lies in the Nihewan Basin, a region in northern China situated between the Chinese Loess Plateau and the Inner Mongolian Plateau. The basin preserves extensive Quaternary fluvio-lacustrine and loess deposits ([@li2000; @yuan2009]), hosts the highest concentrations of Pleistocene Paleolithic sites in East Asia, and is especially significant for preserving evidence of early hominin activities outside Africa dating back over one million years ([@yang2020; @yang2017; @pei2019; @pei2017; @zhu2001; @zhu2004]). In recent years, nearly 20 Middle Pleistocene sites—including Jijiazhuang and Caijiagou—have been identified in the southeastern part of the basin ([@yuwei2023; @ye2024]), within the Yuxian Sub-basin, thereby completing the regional cultural sequence of the Nihewan Basin.

Qianshangying site complex consisting of 4 localities from QSY-A to QSY-D, discovered in 2015, is among the region’s most artifact-rich archaeological sites. Geomorphological and sedimentological evidence indicates that the stratigraphy at the site records a full cycle of lake expansion, retreat, re-expansion, and eventually disappearance of the ancient Nihewan lake, and hominin occupation occurred in a marginal lacustrine setting during a low lake-level episode.

The stratigraphic context of QSY indicates a short-term occupation resulting from a single hominin activity event. The assemblage appears to be largely in situ, exhibiting minimal disturbance after deposition and well-preserved patterns of technological combination. ESR dating of the cultural layer places hominin activity at approximately 429 ± 39 ka.

A total of 957 stone artifacts were recovered from four localities at the QSY site, with QSY-A and QSY-B yielding the most abundant assemblages. Following a widely used techno-typological framework (Isaac, 1981, 1986; de la Torre and Mora, 2005, 2018; de la Torre et al., 2011; Pei et al., 2017), the assemblage was classified into four general categories: flaked pieces (n = 247, 25.9%), detached pieces (n = 686, 71.9%), pounded pieces (n = 4, 0.4%), and unmodified materials (n = 17, 1.8%).

# Method
